---
title: "wateRmelon Vignette"
author: "Connor Smith"
output:
  html_document:
    df_print: paged
---

```{r install}
## try http:// if https:// URLs are not supported
#source("https://bioconductor.org/biocLite.R")
#biocLite("wateRmelon")
library(wateRmelon)
```

# Vignette replication  

## 3 Trying it out  
Load in the data from the *melon* set  
```{r load_data}
# load data
data(melon)
# get dimensions
dim(melon)
# filter quality based on p values - using default from wateRmelon with a threshold of 0.05
# also removes sites w/ 5% of samples having beadcount less than 3 (5% is default)
# also remove sites where 1% of samples have a detection p-value greater than 0.05
melon.pf <- pfilter(melon)
```  

Now pre-process using the best method defined by paper, DaSeN:  
- D: background adjustment performed on Type I and Type II  
- S: Type I and Type II intensities quantile normalized seperately  
- N: no dye normalization performed, methylated and umnethylated normalized seperately  

```{r perform_dasen}
melon.dasen.pf <- dasen(melon.pf)
```

## 4 Performance Metrics  
Notes:  
- all metrics designed so that lower score = better performance  

### 4.1 Genomic Imprinting  
We expect to see hemimethylation in imprinted regions. The first metric looks at these known regions and creates a "standard-error-like measure of dispersion".  

| melon.pf | melon.dasen.pf |
|:--------:|:--------------:|
| `r dmrse_row(melon.pf)` | `r dmrse_row(melon.dasen.pf)` |







